Design and implementaiton of a live coding environment for data science

### 2017-10-25 live/paper.tex
Abstract and introduction

\theoremstyle{plain}
\newtheorem{theorem}[thc]{Theorem}
\newtheorem{lemma}[thc]{Lemma}

@@

\title{Design and implementaiton of a live coding environment}

@@

\begin{document}
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
\end{abstract}
\maketitle

@@

Increasing number of programming environments aim to provide the same live experience for more
standard programming languages, but doing this is not easy. Fully recomputing the whole program after
every single change is inefficient and calculating how a change in source code changes the result
is extremely hard when the editor allows arbitrary manipulation of program text. For example, 
consider the following simple program that gets the release years of 10 most expensive movies 
in a data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}  
\kvd{let}~\ident{top} =\ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\num{10})\\
\quad .\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))}\\
\end{array}
\end{equation*}
%
A live coding environment shows a preview of the list of dates \ident{top} and the programmer then
changes the code by making $\num{10}$ a variable and changing the date format to see the full date:

\begin{equation*}
\begin{array}{l}  
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\ident{count})\\
\quad \ident{.map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))}\\
\end{array}
\end{equation*}
%
Ideally, the live coding environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking 10 elements) and only evaluate the last \ident{map}
and format the release dates of already computed top 10 movies.

This is not difficult to achieve if we represent the program in a structured way and allow the user
to edit code only via well-understood primitive operations such as ``extract variable'' (which has
no effect on the result) or ``change constant value'' (which forces recomputation of subsequent
transformations). However, many programmers still prefer to edit programs as free form text.
In this paper, we present the design and implementation of a live coding system that is capable
of reusing previously evaluated expressions as in the example above, yet, is integrated into an
ordinary text editor.

The motivation for this paper is twofold. First, implementing a live programming system requires
a different way of thinking about compilers and interpreters than the one presented in classic
programming language literature. Second, an increasing number of programming environments 
incorporate aspects of live programming [X,Y], yet relatively little has been written on their
architecture and implementation.

@@

\section{Live programming for data exploration}

@@

\newpage
\section{Properties}

@@

\subsection{Correctness of previews}
Eliminating let bindings does not change the result in the host language. It also yields the same
dependency graph.

\begin{lemma}[Let elimintion]
Given an expression $e_1$ such that $e_1 \rightsquigarrow_\ident{let} e_2$ and a lookup table $\Delta$
then if $\bndclr{v_1}, (V_1, E_1) = \ident{bind}_{\emptyset, \Delta}(e_1)$ and  
$\bndclr{v_2}, (V_2, E_2) = \ident{bind}_{\emptyset, \Delta}(e_2)$ then it holds that
$\bndclr{v_1} = \bndclr{v_2}$ and also $(V_1, E_1) = (V_2, E_2)$.
\end{lemma}

@@

\begin{lemma}[Let-free correctness]
Given an expression $e$ that has no free variables and does not contain let bindnigs, together with
a lookup table $\Delta$ obtained from any sequence of expressions according to Figure~\ref{fig:loop}, 
let $\bndclr{v}, (V, E) = \ident{bind}_{\emptyset, \Delta}(e)$. If $\bndclr{v}\Downarrow d$ 
over a graph $(V, E)$ then $d = p$ for some $p$ and $e \rightsquigarrow p$.
\end{lemma}
\begin{proof}
Every dependency graph node $\bndclr{v}$ corresponds to an expression $e$ with free variables
$FV(e)$. We obtain preview $\bndclr{v}\Downarrow d$ such that if $FV(e) = \emptyset$ then
$d = p$ for some $p$, otherwise, $d = \llbracket e' \rrbracket_{FV(e)}$.  
We show that either $e \rightsquigarrow p$ or that for any $C$ such that $FV(C[e]) = \emptyset$
it holds that $C[e] \rightsquigarrow e''$ and $C[e'] \rightsquigarrow e''$.
By induction over the structure of the evaluation:

@@

\subsection{Reuse of previews}

@@

>>>>

\theoremstyle{plain}
\newtheorem{lem}[thc]{Lemma}
\newtheorem{theorem}[thc]{Theorem}

@@

\title{Design and implementaiton of a live coding environment for data science}

@@

\begin{document}
\begin{abstract}
Data science can be done by directly manipulating data using spreadsheets, or by writing 
data manipulation scripts using a programming language. The former is error-prone and
does not scale, while the latter requires expert skills. Live coding has the potential to 
bridge this gap and make writing of transparent, reproducible scripts more accessible.

In this paper, we describe a live programming environment for data science that provides instant 
previews and contextual hints, while allowing the user to edit code in an unrestricted way in
a text editor. This is challenging as the user can make arbitrary edits and we cannot fully 
recompute previews after every change. The presented interpreter partially evalautes programs as 
they are written and reuses results of previous evaluations, making it possible to provide instant 
feedback during development.

We formalise how programs are interpreted and how previews are computed, prove correctness
of the previews and formally specify when can previews be reused. We hope this work will provide
solid and easy to reuse foundations for the emerging trend of live programming environments.
\end{abstract}
\maketitle

@@

Increasing number of programming environments aim to provide the same live experience for more
standard programming languages, but doing this is not easy. Fully recomputing the whole program after
every single change is inefficient and calculating how a change in source code changes the result
is extremely hard when the editor allows arbitrary manipulation of program text. For example, 
consider the following simple program that gets the release years of 10 most expensive movies 
in a data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}  
\kvd{let}~\ident{top} =\ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\num{10})\\
\quad .\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))}\\
\end{array}
\end{equation*}
%
A live coding environment shows a preview of the list of dates \ident{top} and the programmer then
modifies the code by making the constant $\num{10}$ a variable and changing the date format to see 
the full date:
%
\begin{equation*}
\begin{array}{l}  
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\ident{count})\\
\quad \ident{.map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))}\\
\end{array}
\end{equation*}
%
Ideally, the live coding environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking 10 elements) and only evaluate the last \ident{map}
to differently format the release dates of already computed top 10 movies.

This is not difficult to achieve if we represent the program in a structured way and allow the user
to edit code only via well-understood primitive operations such as ``extract variable'' (which has
no effect on the result) or ``change constant value'' (which forces recomputation of subsequent
transformations). However, many programmers prefer to edit programs as free form text.
We present the design and implementation of a live coding system that is capable
of reusing previously evaluated expressions as in the example above, yet, is integrated into an
ordinary text editor. The main contributions of this paper are:

\begin{itemize}
\item We introduce The Gamma (Section~\ref{sec:live}), a simple live coding environment for
  data science. We review its design and implementation and explain how it 
  bridges the gap between programming and spreadsheets.

\item Implementing a live programming system requires different way of thinking about compilers
  and interpreters than the one presented in classic programming language literature. Our 
  formalisation (Section~\ref{sec:formal}) captures the essence of the new perspective.

\item We formalise the evaluation of previews (Section~\ref{sec:previews}) and prove that our
  evaluation and caching mechanism is produces correct previews (Section~\ref{sec:properties-correct}) 
  and can effectively reuse partial results (Section~\ref{sec:properties-reuse}).

\item In the final more speculative part (Section~\ref{sec:design}), we consider alternative
  language designs that would enable further live coding experiences, which are difficult to
  build using our current system.
\end{itemize}

\noindent
We hope the architecture and its formal presentation in this paper can contribute important 
foundations to the growing and important trend of text-based live coding environments.

@@

\section{Live programming for data exploration}
\label{sec:live}

@@

\newpage
\section{Properties}
\label{sec:properties}

@@

\subsection{Correctness of previews}
\label{sec:properties-correct}

Eliminating let bindings does not change the result in the host language. It also yields the same
dependency graph.

\begin{lem}[Let elimintion]
Given an expression $e_1$ such that $e_1 \rightsquigarrow_\ident{let} e_2$ and a lookup table $\Delta$
then if $\bndclr{v_1}, (V_1, E_1) = \ident{bind}_{\emptyset, \Delta}(e_1)$ and  
$\bndclr{v_2}, (V_2, E_2) = \ident{bind}_{\emptyset, \Delta}(e_2)$ then it holds that
$\bndclr{v_1} = \bndclr{v_2}$ and also $(V_1, E_1) = (V_2, E_2)$.
\end{lem}

@@

\begin{lem}[Let-free correctness]
Given an expression $e$ that has no free variables and does not contain let bindnigs, together with
a lookup table $\Delta$ obtained from any sequence of expressions according to Figure~\ref{fig:loop}, 
let $\bndclr{v}, (V, E) = \ident{bind}_{\emptyset, \Delta}(e)$. If $\bndclr{v}\Downarrow d$ 
over a graph $(V, E)$ then $d = p$ for some $p$ and $e \rightsquigarrow p$.
\end{lem}
\begin{proof}
Every dependency graph node $\bndclr{v}$ corresponds to an expression $e$ with free variables
$FV(e)$. We obtain preview $\bndclr{v}\Downarrow d$ such that if $FV(e) = \emptyset$ then
$d = p$ for some $p$, otherwise, $d = \llbracket e' \rrbracket_{FV(e)}$.  
We show that either $e \rightsquigarrow p$ or that for any $C$ such that $FV(C[e]) = \emptyset$
it holds that $C[e] \rightsquigarrow e''$ and $C[e'] \rightsquigarrow e''$.
By induction over the structure of the evaluation:

@@

\subsection{Reuse of previews}
\label{sec:properties-reuse}

@@

\section{Takeaways}
\label{sec:design}
